---
title: "Covid19"
author: "Brandon Khuu"
date: "2024-06-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(purrr)
library(caret)
library(randomForest)
```

Import the data
```{r}
# github link
url_in = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

#  the csv files
file_names = c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_confirmed_US.csv")

# Import data
urls = str_c(url_in, file_names)
urls
```

Print Overview of each table

```{r}
us_cases = read_csv(urls[3])
us_deaths = read_csv(urls[4])
```
Check the First 5 of each table
```{r}
head(us_cases)
head(us_deaths)
```

Preprocess and clean the tables for the columns that we will be working with

```{r}
# Pivot wide-format data for dates and sum totals for each state

cases = us_cases %>% 
  pivot_longer(cols = 13:ncol(us_cases), names_to = "date") %>%
  group_by(`Province_State`, date) %>%
  summarise("cum_cases" = sum(value, na.rm = TRUE), .groups = 'drop')

deaths = us_deaths %>% 
  pivot_longer(cols = 13:ncol(us_deaths), names_to = "date") %>%
  group_by(`Province_State`, date) %>%
  summarise("cum_deaths" = sum(value, na.rm = TRUE), .groups = 'drop')
```

Adjust the date from character string to date object
```{r}
# Convert dates to datetime object
cases$date = lubridate::mdy(cases$date)
deaths$date = lubridate::mdy(deaths$date)

# Rename columns from Province_State -> State & Admin2 -> County
cases = cases %>% 
  rename_at('Province_State', ~'State')
deaths = deaths %>% 
  rename_at('Province_State', ~'State')

# check global deaths and cases data 
head(cases)
head(deaths)
```
Look for the states with the highest number of caeses and deaths to do analysis
```{r}

top_cases <- cases %>%
  group_by(State) %>%
  summarise(cum_cases = max(cum_cases, na.rm = TRUE))

top_deaths <- deaths %>%
  group_by(State) %>%
  summarise(cum_deaths = max(cum_deaths, na.rm = TRUE))

top5cases <- top_cases %>%
  arrange(desc(cum_cases)) %>%
  slice_head(n=5)

top5deaths <- top_deaths %>%
  arrange(desc(cum_deaths)) %>%
  slice_head(n=5)

print(top5cases)
print(top5deaths)
```

Create us_all and the split by target states
```{r merge global cases and deaths}
# merge global data sets and filter to get data just for Switzerland and for germany
us_all = merge(x=deaths, y=cases, all.x=TRUE)
CA <- us_all[us_all$State == "California", ]
TX <- us_all[us_all$State == "Texas", ]
FL <- us_all[us_all$State == "Florida", ]
NY <- us_all[us_all$State == "New York", ]
IL <- us_all[us_all$State == "Illinois", ]
PA <- us_all[us_all$State == "Pennsylvania", ]

# View first several lines of each data set

head(CA)
head(TX)
head(FL)
head(NY)
head(IL)
head(PA)
```

Adjust the values in the columns based on state popululation. (per capita)
[State populations from: [https://worldpopulationreview.com/states]
```{r}

# Adding a per capita column to both data sets and viewing their summaries
CA_cap <- CA %>%
  mutate(
    CA_cases_per_capita = cum_cases / 38889770,
    CA_deaths_per_capita = cum_deaths / 38889770,
  )


TX_cap <- TX %>%
  mutate(
    TX_cases_per_capita = cum_cases / 30976754,
    TX_deaths_per_capita = cum_deaths / 30976754
  )

FL_cap <- FL %>%
  mutate(
    FL_cases_per_capita = cum_cases / 22975931,
    FL_deaths_per_capita = cum_deaths / 22975931
  )


NY_cap <- NY %>%
  mutate(
    NY_cases_per_capita = cum_cases / 19469232,
    NY_deaths_per_capita = cum_deaths / 19469232
  )

IL_cap <- IL %>%
  mutate(
    IL_cases_per_capita = cum_cases / 12516863,
    IL_deaths_per_capita = cum_deaths / 12516863
  )

PA_cap <- PA %>%
  mutate(
    PA_cases_per_capita = cum_cases / 12516863,
    PA_deaths_per_capita = cum_deaths / 12516863
  )
```

Print smmary of each state to get an overview of each dataframe
```{r}
summary(CA_cap)
summary(TX_cap)
summary(FL_cap)
summary(NY_cap)
summary(IL_cap)
summary(PA_cap)
```

Merge the individual state dataframes into 1 single dataframe
```{r}
merged_CA_TX <- merge(CA_cap,TX_cap, by = "date", all = TRUE, suffixes = c(".CA", ".TX"))
merged_CA_TX_FL <- merge(merged_CA_TX,FL_cap, by = "date", all = TRUE, suffixes = c("", "FL"))
merged_CA_TX_FL_NY <- merge(merged_CA_TX_FL,NY_cap, by = "date", all = TRUE, suffixes = c("", "NY"))
merged_CA_TX_FL_NY_PA <- merge(merged_CA_TX_FL_NY,PA_cap, by = "date", all = TRUE, suffixes = c("", "PA"))
all_merge <- merge(merged_CA_TX_FL_NY_PA,IL_cap, by = "date", all = TRUE, suffixes = c("", "IL"))
```

Plot the cases per capita and deaths per capita
```{r}
# Create line plots for cases per capita
ggplot(all_merge, aes(x = date)) +
  geom_line(aes(y = CA_cases_per_capita, color = "California", linetype = "California")) +
  geom_line(aes(y = TX_cases_per_capita, color = "Texas", linetype = "Texas")) +
  geom_line(aes(y = FL_cases_per_capita, color = "Florida", linetype = "Florida")) +
  geom_line(aes(y = NY_cases_per_capita, color = "New York", linetype = "New York")) +
  geom_line(aes(y = IL_cases_per_capita, color = "Illinois", linetype = "Illinois")) +
  geom_line(aes(y = PA_cases_per_capita, color = "Pennsylvania", linetype = "Pennsylvania")) +
  labs(title = "Cumulative Cases Per Capita Comparison",
       x = "Date",
       y = "Cumulative Cases Per Capita",
       color = "State",
       linetype = "State") +
  scale_color_manual(values = c("California" = "blue", 
                                "Texas" = "red",
                                "Florida" = "green", 
                                "New York" = "purple", 
                                "Illinois" = "pink",
                                "Pennsylvania" = "yellow")) +
  scale_linetype_manual(values = c("California" = "solid", 
                                   "Texas" = "solid", 
                                   "Florida" = "solid", 
                                   "New York" = "solid", 
                                   "Illinois" = "solid",
                                   "Pennsylvania" = "solid")) +
  theme_minimal()

# Create line plots for deaths per capita
ggplot(all_merge, aes(x = date)) +
  geom_line(aes(y = CA_deaths_per_capita, color = "California", linetype = "California")) +
  geom_line(aes(y = TX_deaths_per_capita, color = "Texas", linetype = "Texas")) +
  geom_line(aes(y = FL_deaths_per_capita, color = "Florida", linetype = "Florida")) +
  geom_line(aes(y = NY_deaths_per_capita, color = "New York", linetype = "New York")) +
  geom_line(aes(y = IL_deaths_per_capita, color = "Illinois", linetype = "Illinois")) +
  geom_line(aes(y = PA_deaths_per_capita, color = "Pennsylvania", linetype = "Pennsylvania")) +
  labs(title = "Cumulative Deaths Per Capita Comparison",
       x = "Date",
       y = "Cumulative Deaths Per Capita",
       color = "State",
       linetype = "State") +
  scale_color_manual(values = c("California" = "blue", 
                                "Texas" = "red",
                                "Florida" = "green", 
                                "New York" = "purple", 
                                "Illinois" = "pink",
                                "Pennsylvania" = "yellow")) +
  scale_linetype_manual(values = c("California" = "solid", 
                                   "Texas" = "solid", 
                                   "Florida" = "solid", 
                                   "New York" = "solid", 
                                   "Illinois" = "solid",
                                   "Pennsylvania" = "solid")) +
  theme_minimal()
```
Create columnn deaths per case
```{r}
CA$deaths_per_case <- ifelse(CA$cum_cases > 0,CA$cum_deaths / CA$cum_cases,0)
TX$deaths_per_case <- ifelse(TX$cum_cases > 0,TX$cum_deaths / TX$cum_cases,0)
FL$deaths_per_case <- ifelse(FL$cum_cases > 0,FL$cum_deaths / FL$cum_cases,0)
NY$deaths_per_case <- ifelse(NY$cum_cases > 0,NY$cum_deaths / NY$cum_cases,0)
IL$deaths_per_case <- ifelse(IL$cum_cases > 0,IL$cum_deaths / IL$cum_cases,0)
PA$deaths_per_case <- ifelse(PA$cum_cases > 0,PA$cum_deaths / PA$cum_cases,0)
```
Plot deaths per case to view deaths over time
```{r}

ggplot(CA, aes(x = date, y = deaths_per_case, color = "California")) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "California") +
  theme_minimal()

ggplot(TX, aes(x = date, y = deaths_per_case, color = "Texas")) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "Texas") +
  theme_minimal()

ggplot(FL, aes(x = date, y = deaths_per_case, color = "Florida")) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "California") +
  theme_minimal()

ggplot(NY, aes(x = date, y = deaths_per_case, color = "New York")) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "New York") +
  theme_minimal()

ggplot(IL, aes(x = date, y = deaths_per_case, color = "Illinois")) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "Illinois") +
  theme_minimal()

ggplot(PA, aes(x = date, y = deaths_per_case, color = "Pennsylvania")) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "Illinois") +
  theme_minimal()

```
Merge each state dataframe to view on a single graph
```{r}
merged_states_count <- rbind(transform(CA, dataset = "CA"),
                             transform(TX, dataset = "TX"),
                             transform(FL, dataset = "FL"),
                             transform(NY, dataset = "NY"),
                             transform(IL, dataset = "IL"),
                             transform(PA, dataset = "PA"))

ggplot(merged_states_count, aes(x = date, y = deaths_per_case, color = State, linetype = dataset)) +
  geom_line() +
  labs(title = "Deaths per Case Over Time",
       x = "Date",
       y = "Deaths per Case",
       color = "State",
       linetype = "Dataset") +
  theme_minimal()
```

State predictor model with random forest.
Principal Compnent Analysis(PCA) was used as the target feature to aggregate all 6 deaths per capita to maintain the variability 
```{r}

X <- all_merge[, c('CA_cases_per_capita', 'TX_cases_per_capita', 
                   'FL_cases_per_capita', 'NY_cases_per_capita', 
                   'IL_cases_per_capita', 'PA_cases_per_capita')]

pca <- prcomp(all_merge[, c('CA_deaths_per_capita', 'TX_deaths_per_capita', 
                            'FL_deaths_per_capita', 'NY_deaths_per_capita', 
                            'IL_deaths_per_capita','PA_deaths_per_capita')], scale = TRUE)

all_merge$pca_deaths_per_capita <- pca$x[,1]
y <- all_merge$pca_deaths_per_capita
```


Split data into train/test set using seed 123
```{r}
set.seed(123)
trainIndex <- createDataPartition(y, p = .8, list = FALSE, times = 1)
X_train <- X[trainIndex,]
X_test <- X[-trainIndex,]
y_train <- y[trainIndex]
y_test <- y[-trainIndex]
rf_model <- randomForest(X_train, y_train, ntree = 100)
```

Predictions form the model and scoring with Mean Squarred Error(MSE)
```{r}
y_pred <- predict(rf_model, X_test)
mse <- mean((y_test - y_pred)^2)
print(paste('Mean Squared Error:', mse))
print(importance(rf_model))
```

Graph Node Purity to see which column had the most influnce on the model
```{r}
importance_df <- as.data.frame(importance(rf_model))
importance_df$Feature <- rownames(importance_df)

# Create a bar plot
ggplot(importance_df, aes(x = reorder(Feature, IncNodePurity), y = IncNodePurity)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Feature") +
  ylab("Increase in Node Purity") +
  ggtitle("Feature Importance based on IncNodePurity") +
  theme_minimal()

```

