---
title: "NYPD_Shooting_Data_History"
author: "Brandon Khuu"
date: "2024-03-02"
output: html_document
---

#Import Libraries

Import the libraries needed to run the file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(caret)
library(randomForest)
```


#Import the file

Import the library in R studio. Add a header and do not convert the strings into a factor data type

```{r }
data1 <- read.csv("C:\\Users\\bkhuu\\ucb\\NYPD_Shooting_Incident_Data__Historic_.csv", header=TRUE, stringsAsFactors = FALSE)

```

## Data Summary

Initial data summary to describe the data from a surface level view.

```{r }
str(data1)
summary(data1)
```

## Check for missing values

See which (if any) columns have missing values and if there are any things we need to impute for analysis and models going forward.

Depending on the columns and values we can decide on the imputation method.

```{r }

colSums(is.na(data1))
```

## Convert dates to Date format

Convert the date format into something a little easier to work with
```{r }

data1$OCCUR_DATE <- as.Date(data1$OCCUR_DATE, "%m/%d/%Y")
```

##Exploratory analysis

Do some basic analysis and visualization via graphs

```{r}
data1 %>%
  mutate(Year = format(OCCUR_DATE, "%Y")) %>%
  group_by(Year) %>%
  summarise(Incidents = n()) %>%
  ggplot(aes(x = Year, y = Incidents)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Number of Shooting Incidents per Year",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal()

```
The graph above shows the number of shootings per years

```{r}
data1 %>%
  group_by(BORO) %>%
  summarise(Incidents = n()) %>%
  ggplot(aes(x = BORO, y = Incidents, fill = BORO)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Shooting Incidents by Borough",
       x = "Borough",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(legend.position = "none")
```

This graph shows the number of shootings per boro.

#Preparing data for the model

##set historical dates
creating a filter for older dates to be used in the future
```{r}

# Filter data to include only incidents before the specified date
data_before_date <- filter(data1, OCCUR_DATE < as.Date("2024-03-05"))

# Create a new column for the day of the year using as.numeric and format instead of yday
data_before_date$DayOfYear <- as.numeric(format(data_before_date$OCCUR_DATE, "%j"))

```

Splitting data into training sets and testing sets.
Setting seed to "123" which will allow for reproducibility.

```{r}
# Splitting the data into training and testing sets
set.seed(123)
training_indices <- createDataPartition(data_before_date$DayOfYear, p=0.8, list=FALSE)
training_data <- data_before_date[training_indices,]
testing_data <- data_before_date[-training_indices,]
```

Train the model
```{r}
model <- randomForest(DayOfYear ~ ., data=training_data[,-which(names(training_data) %in% c("OCCUR_DATE"))], ntree=100, na.action = na.omit)
```

Test the model
```{r}
# Calculating the Mean Squared Error (MSE) for the model's predictions, ensuring no NA values are included in the calculation
actual_days <- testing_data$DayOfYear
predicted_days <- predict(model, newdata=testing_data[,-which(names(testing_data) %in% c("OCCUR_DATE", "DayOfYear"))])

# Filtering out any NA values from actual_days and predicted_days before calculating MSE
valid_indices <- !is.na(actual_days) & !is.na(predicted_days)
mse <- mean((actual_days[valid_indices] - predicted_days[valid_indices])^2, na.rm = TRUE)
print(paste("Mean Squared Error (MSE):", mse))
```
Accuracy of the model was scored using Mean Squared Error.
MSE: 5761.17499135467

```{r}
# Ensuring the prediction is a numeric value and handling potential NA values
current_year_incident_prediction <- na.omit(predict(model, newdata=testing_data[,-which(names(testing_data) %in% c("OCCUR_DATE", "DayOfYear"))]))
if(length(current_year_incident_prediction) > 0) {
  # Extracting the current year from the system's date
  current_year <- format(Sys.Date(), "%Y")
  # Converting the mean of predictions to a future date
  predicted_future_date <- as.Date(paste(current_year, "-01-01", sep="")) + mean(current_year_incident_prediction) - 1
  print(paste("Predicted future date of the next incident:", as.character(predicted_future_date)))
} else {
  print("Unable to predict future date due to insufficient data or model prediction error.")
}
```
Predicted future date of the next incident using the testing data: 2024-07-06